---
title: "Colors"
author: "Me"
date: "7/31/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Setting the colors of some geoms



```{r}
library(tidyverse)
library(ggplot2)
library(patchwork)
library(gcookbook)
library(MASS)
```
In the call geom(), set the values to color or fill

```{r}
a = ggplot(mtcars, aes(x = wt, y = mpg)) + geom_point(color = 'red')
b = ggplot(birthwt, aes(x = bwt)) + geom_histogram(fill = 'red', color = 'black')

a + b
```


### Mapping variables to color

```{r}
a = ggplot(cabbage_exp, aes(x = Date, y = Weight, fill = Cultivar)) + geom_col(color = 'black', position = 'dodge')

b = ggplot(mtcars, aes(x = wt, y = mpg, color = cyl)) + geom_point()

a + b
```
Set the cyl to factor to make it a discrete variable

```{r}
b = ggplot(mtcars, aes(x = wt, y = mpg, color = cyl)) + geom_point()

c = ggplot(mtcars, aes(x = wt, y = mpg, color = factor(cyl))) + geom_point()

b +c 
```


### Using a different palette 

Use one of the scales listed
scale_color_discrete - evenly spaced color
scale_color_hue - same as discrete
scale_color_grey - greyscale palette
scale_color_brewer - colorbrewer pallets
scale_color_manual - manually specified collors


```{r}
p = ggplot(uspopage, aes(x = Year, y = Thousands, fill = AgeGroup)) + geom_area()

p + scale_fill_discrete() + p + scale_fill_brewer()
```

There are two types of scales that use color: fill scales and color scales

The default lightness value is 65. This is good for filled areas but a little light on points and lines.

```{r}
h = ggplot(heightweight, aes(x = ageYear, y = heightIn, color = sex))+
  geom_point()

h + h + scale_color_hue(l = 30)
```

Scale color brewer provides a number of palettes
```{r}
library(RColorBrewer)
display.brewer.all()
```

```{r}
p + scale_fill_brewer(palette = 'Oranges')
```
You can also use a paletter of greys. The default start is .2 and default end is .8

```{r}
p + scale_fill_grey(start = .7, end = 0)
```

### Using a manually defined palette
Specify the colors in scale_color_manual()

```{r}
h = ggplot(heightweight, aes(x = ageYear, y = heightIn, color = sex)) + geom_point()

h + scale_color_manual(values = c('red','blue'))
```
### Using a manually defined palette for a continuous variable

```{r}
library(scales)

p = ggplot(heightweight, aes(x = ageYear, y = heightIn, color = weightLb)) + geom_point(size = 3)

#gradient between two colors
a = p + scale_color_gradient(low = 'black', high = 'white')
#with a white midpoint
b = p + scale_color_gradient2(low = muted('red'), mid = 'white', high = muted('blue'))
#gradient n colors
c = p + scale_color_gradientn(colors = c('darkred','orange','yellow','white'))

a/b/c
```

## color a shaded region base on value

```{r}
cb = subset(climate, Source == 'Berkeley')

cb %>%
  mutate(valence = ifelse(Anomaly10y >= 0, 'pos', 'neg')) %>%
  ggplot(aes(x = Year, y = Anomaly10y))+
  geom_line()+
  geom_area(aes(fill = valence)) +
  geom_hline(yintercept = 0, color = 'white', size = rel(1.2))+
  scale_fill_manual(values = c('#CCEEFF','#FFDDDD'), guide = F)+
  scale_x_continuous(expand = c(0,0))
```





















