---
title: "Facets"
author: "Me"
date: "7/31/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### Splitting data into subplots with facets

use facet_grid() or facet_wrap() and specify which variable to split


```{r}
library(ggplot2)
library(gcookbook)
library(tidyverse)
library(patchwork)

#base plot

p = ggplot(mpg, aes(x = displ, y = hwy)) + geom_point()

#vertically arranged panels
a = p + facet_grid(drv ~ .)

#horizontally arrange panels
b = p + facet_grid(.~ drv)

a + b
```
SPlit by vertical and horizontal
```{r}
p + facet_grid(drv ~ cyl)
```

With facet_wrap(), subplots are laid out horizontally and wrap around like words on a page

```{r}
p + facet_wrap( ~ class)
```

number of rows and columns could be specified in facet

```{r}
p + facet_wrap(~class, nrow = 2)
```


### Using facets with different axes

set the scales to free_x or free_y

```{r}
p + facet_grid(drv ~ cyl, scales = 'free')
```

### Changing the 

```{r}
mpg2 = mpg

#name hte levels

mpg2 %>%
  mutate(drv = case_when(
    drv == 'f' ~ 'front drive',
    drv == '4' ~ 'forward',
    drv == 'r' ~ 'rear'
  )) %>%
  ggplot(aes(displ, hwy)) +
  geom_point()+
  facet_wrap(~drv)
```
### Changing the apperance of facet labels and headers

use the strip in theme()

```{r}
ggplot(cabbage_exp, aes(x = Cultivar, y = Weight)) +
  geom_col()+
  facet_wrap(~ Date) +
  theme(
    strip.text = element_text(face = 'bold', size = rel(1.5)),
    strip.background = element_rect(fill = 'lightblue', color = 'black', size = 1)
    
  )
```



























