---
title: "Legends"
author: "Me"
date: "7/31/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### Removing the legend

Use guides() and specify the scale that should have its legend removed

```{r}
library(ggplot2)
library(tidyverse)
library(gcookbook)
library(patchwork)

p = ggplot(PlantGrowth, aes(x  = group, y = weight, fill = group)) + geom_boxplot()

a = p + guides(fill = F)

b = p + scale_fill_discrete(guide = F)

a + b
```
You can also remove it from the theme

p + theme(legend.position = 'none')

when a variable is mapped to fill, the default scale used is scale_fill_discrete() which maps the factors levels to colors that are equally spaced around the color wheel. There are other scales for fill such as scale_fill_manual(). if you use scales for other aesthetics such as color or shape, you must use the appropriate scale. Commonly used scales are

- scale_fill_discrete
- scale_fill_hue
- scale_fill_manual
- scale_fill_grey
- scale_fill_brewer
- scale_color_discrete
- scale_color_hue
- scale_color_manual
_ scale_color_brewer
_ scale_shape_manual
- scale_linetype

### Changing the position of the legend

use theme(legend.position = ...)

```{r}
p = ggplot(PlantGrowth, aes(x = group, y = weight, fill = group)) +
  geom_boxplot() + 
  scale_fill_brewer(palette = 'Pastel2')

p + theme(legend.position = 'top')
```


legend can also be placed inside the graphing area by coordinate (0,0) is the bottom left and goes to (1,1) on the upper right

```{r}
a = p + theme(legend.position = c(1,0), legend.justification = c(1,0))

b = p + theme(legend.position = c(1,1), legend.justification = c(1,1))

a + b
```

When placing the legend inside the graphing area, it may be helpful to add opaque border to it

```{r}
p + theme(legend.position = c(.85, .2))+
  theme(legend.background = element_rect(fill = 'white', color = 'black'))+
  theme(legend.key = element_blank())
```


### Changing the order of items in a legend

set the limits to the desired order

```{r}


p + p + scale_fill_discrete(limits = c('trt1','trt2','ctrl'))
```

Note that the order of items on the x axis did not change. To do that you need to have set the limits in the scale_x_discrete()

In the preceding example, the group was mapped to fill aesthetic. By default this uses scale_fill_discrete(), which maps factor levels to colors that are equally spaced around the color wheel. We could have used a different scale_fill_xxx(), though. 

```{r}
a = p + scale_fill_grey(start = .5, end = 1, limits = c('trt1','trt2','ctrl'))

b = p + scale_fill_brewer(palette = 'Pastel2',  limits = c('trt1','trt2','ctrl'))

a + b
```

All the previous examples were for fill. If you use scales for other aesthetics such as color or shape, you must use the appropriate scale

commonly used scales include:
- scale_fill_discrete
- scale_fill_hue
- scale_fill_manual
- scale_fill_grey
- scale_fill_brewer
- scale_color_discrete
- scale_color_hue
- scale_Color_manual
- scale_color_grey
- scale_color_brewer
- scale_shape_manual
- scale_linetype

### Reversing the order of items in a legend

add guides(fill = guide_legend(reverse = T))

```{r}

p + p + guides(fill = guide_legend(reverse = T))
```

It is also possible to control the legend when specifying the scale

scale_fill_hue(guide = guide_legend(reverse = T))


### Changing a legend title

use labs() and set the value of fill, color, shape or whatever aesthetic is appropriate


```{r}
a = p + labs(fill = 'condition')

a + p + scale_fill_discrete(name = 'condition')
```

If there are multiple variables mapped to aesthetics, you can set the title individually

```{r}
hw = ggplot(heightweight, aes(x = ageYear, y = heightIn, color = sex))+
  geom_point(aes(size = weightLb)) +
  scale_size_continuous(range = c(1,4))

hw +
  labs(color = 'male/Female',
       size = 'Weight\n(pounds)')
```
If a variable is mapped to both shape and color. you msut rename bothh

```{r}
hw1 = ggplot(heightweight, aes(x = ageYear, y = heightIn, shape = sex, color = sex)) + geom_point()

a = hw1 + 
  labs(shape = 'Male/Female', color = 'Male/Female')

b = hw1 +
  guides(fill = guide_legend(title = 'Male/Female'))

a + b
```
### Changing the appearance of legend title

use theme(legend.title = element_text(...))

```{r}
p +
  theme(
    legend.title = element_text(face = 'italic', family = 'serif', color = 'red', size = 14)
  )
```
### Removing a title legend

add guides(fill = guide_legend(title = NULL))

```{r}
a = p +
  guides(fill = guide_legend(title = NULL))

b = p +
  labs(fill = '')

a + b
```

### Changing labels in the legend

Set the labels in the scale

```{r}
p +
  scale_fill_discrete(labels = c('Control','Treatment 1','Treatment 2'))
```

note that the labels on the x axis did not change. To do that you need to change the scale_x_discrete. 

To change the order you can set it in the limits

```{r}
p + scale_fill_discrete(limits = c('trt1','trt2','ctrl'),
                        labels = c('Treatment 1','Treatment 2','Control'))
```


If you have one variable that is mapped to two separate aesthetics. the default is to have a single legend that combines both. if you want to change the labels, you must change them in both scales. 

```{r}
p = ggplot(heightweight, aes(x=  ageYear, y = heightIn, shape = sex, color = sex))+ geom_point()

p + scale_shape_discrete(labels = c('Female','Male'))+
  scale_color_discrete(labels = c('Female','Male'))
```


### Changing the apperance of legend labels


use theme(legend.text = element_text(...))

```{r}
p = ggplot(PlantGrowth, aes(x = group, y = weight, fill = group)) + geom_boxplot()

p + theme(legend.text = element_text(face = 'italic', family = 'serif', color = 'red', size = 14))
```

### Using labels with multiple lines of text


use '\n' to represent a new line. 

```{r}
p + scale_fill_discrete(
  labels = c('Control', 'Type1\nTreatment','Type 2\nTreatment')
)
```

THe lines will run to each other, to deal with this problem, we can increase the height of legend keys and decrease the spacing between them

```{r}
p + scale_fill_discrete(
  labels = c('Control', 'Type1\nTreatment','Type 2\nTreatment')
)+
  theme(legend.text = element_text(lineheight = .8),
        legend.key.height = unit(1.5,'cm'))
```












