---
title: "Build plot layer by layer"
author: "Me"
date: "7/22/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Layering is the mechanism by which additional data elements are added to a plot. 


Creating a plot - ggplot has two arguments: data and aesthetic mappings. 

```{r}
p = ggplot(diamonds, aes(carat, price, color = cut))
```

Layers - minimal layer do nothing more than specify a geom. A way of visually representing the data

```{r}
p = ggplot(diamonds, aes(x = carat))
p + layer(
  geom = 'bar',
  geom_params = list(fill = 'steelblue'),
  stat = 'bin',
  stat_params = list(binwidth = 2)
)
```
shortcut functions begins with geom_ or stat_

geom_xxx(mapping, data, .... geom, position)
stat_xxx(mapping, data, .... geom, position)


```{r}
ggplot(msleep, aes(sleep_rem/ sleep_total, awake))+
  geom_point()+
  geom_smooth()
```


Plot and layers could be stored as variables. Making it easy to write code for duplication. 


Data - restriction is that the data must be a dataframe. 

Aesthetic mappings - describe the way variables in the data are mapped to things that we can see on the plot. 

Plot and layers - default mappings can be extended or overriden in the layers. Aesthetic mapping specified in a layer only affects that layer. 

Setting vs mapping - we map aesthetic to a variable or set it constant
aes(color = cut)
color = 'red'


Grouping - group is set to the interaction of all discrete variables in the plot. This often partitions data correctly. 

Multiple groups, one aesthetic - in many situations you want to separate your data into groups but render them the same way. 
ggplot(oxboys, aes(age, height, group = subject))+
geom_line()

Different groups on different layers - sometimes we want to plot summaries based on different levels of aggregation. Different layers might have different group aesthetics. 

For example, if we add geom_smooth to all groups, they will have individual geom_smooth. instead you must set group = 1

Overriding default grouping - 

Geoms - each geom has default statistics and every statistics has default geoms. 

stat - transforms the data, typically by summarising it in some manner. 

```{r}
ggplot(diamonds, aes(carat))+
  geom_histogram(aes(y = ..density..), binwidth = .1)
```

The names of generated variables must be surrounded with .. when used to prevent confusion in case the original variable includes the same name. 

Position adjustments - 
Dodge - adjust position by dodging overlaps
fill - stack overlapping objects
identity - dont adjust points
jitter - jitter points to avoid overplotting
stack - stack overlapping points on top of one another


Combining geom and stats - by connecting geoms with different statistics, you can easily create new graphics. The following use same statistical transformation but use different geoms to display the results: area geom, point geom, and tile geom

```{r}
d = ggplot(diamonds, aes(carat))+xlim(0,3)

d+ stat_bin(aes(ymax = ..count..), binwidth = .1, geom = 'area')
```


```{r}
d + stat_bin(
  aes(size= ..density..), binwidth = .1, geom = 'point', position = 'identity'
)
```


Displaying precomputed statistics - if you have data which already has been summarised, youll need to use stat_identity. 





















