---
title: "Chapter 3 - Mastering the grammar"
author: "Me"
date: "7/22/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(ggplot2)
mpg
```


Building a scatterplot - 

```{r}
ggplot(mpg, aes(displ, hwy, color = cyl))+
  geom_point()
```


Scaling - conversion of data units to something a computer can display. 

More complex plot - adds components to the fix : facets, multiple layers, statistics. 
- facets : each facet in each own layer has its own dataseet
- smooth : does not display the raw data but statistical transformation of the data. Other useful stats include 2d binning, group means, quantile regression, and contouring


Components of layered grammar 
- data
- mappings
- geoms
- statistical transformation
- position - deals with overlapping objects. 

Togethere, these forms a layer. A plot may have multiple layers. Altogether a layered grammar defines a plot as the combination of
- Default dataset and set of mappings from one variable to aesthetics
- One or more layers, each of a composed geometric object, statistical transformation, a position adjustment, and optionally a dataset and aesthetic mappings
- one scale for each mapping
- A coordinate system
- A faceting specification


Layers 
- responsible for creating objects we perceive on the plot
- composed of four parts: data and aesthetic mapping, statistical transformation, a geometric object, position adjustmnet

Scales
- controls mapping from data to aesthetic attributes. We need scale for every aesthetic used on plot. 

Coordinate system
- maps the position of objects onto plane of the plot. Position is often specified by two coordinates(x,y) but potentially 3 or more. 
- coordinate system control how axes and grid lines are drawn. 

Faceting
- create small multiples each showing a different subset of whole dataset. This is a powerful tool when investigating whether pattern holds across all conditions. 

Data structures
- A plot object is a list with components data, mapping, layers, scales, coordinates, facets. 
- Once you have a plot object there are few things you can do with it. Render it on screen, render it to a disk with ggsave. describe its structure with summary. save a cached copy with save.


























