---
title: "Polishing your files for publication"
author: "Me"
date: "7/25/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Themes - Themes dont change the perpetual properties of the plot but they do help you make the plot aesthetically pleasing. Themes give you control over things like fonts, titles. axis labels, axis tick labels, strips, legends, labels, legend key labels, color of ticks, grid lines, and background. 

Unlike many other areas in ggplot2, themes can be controlled on multiple levels from the coarse to fine. You can:
- Use built in theme, as described in section 8.1.1. This affects every element of the plot in visually consistent manner. 
- Modify a single element of the built in theme. Each theme is madeup of multiple elements. The theme system comes with a number of built in rendering functions with a limited set of parameters. By adjusting these parameters you can control things like text, size, background, and color. 


Built in themes - 

You can apply apply themes globally affecting all drawn themes using theme_set(theme_grey())
Locally, for individual plot by qplot() + theme_grey()


Theme elements and element functions - a theme is madeup of multiple elements which control the appearance of single item on the plot. There are three elements that have individual x and y settings: axis.text, axis.title, and strip.text. Having a different setting for horizontal and vertical elements allow you to control how text should appear in different orientations. there are four basic types of built in element functions: text, lines, segments, rectangles, and blanks. 

- theme_text() draws labels and headings. You can control the font family, face, color, size, hjust, vjust, angle, and lineheight. 

```{r}
d2 = subset(diamonds, price < 10000)
plot = ggplot(d2, aes(price))+ 
  geom_histogram()

plot +
  labs(title = 'Title: This is a histogram')+
  theme(plot.title = element_text(size = 20, face = 'bold'))


```


- theme_line() and theme_segment() draw lines and segments with the same options but slighlty different way. make sure you match the appropriate type or you will get strange grid errors. 

```{r}
plot +
  theme(panel.grid.major = element_line(color = 'red'))

plot + theme(panel.grid.major = element_line(size = 2))

plot + theme(panel.grid.major = element_line(linetype = 'dotted', size = 2))

plot + theme(axis.line = element_line(color = 'red'))

plot + theme(axis.line = element_line(linetype = 'dashed'))
```

theme_recrt = draws rectangles mostly for backgrounds. You can controll the fill, color size, and linetype.

```{r}
plot + theme(plot.background = element_rect(fill = 'white', color = NA))

plot + theme(plot.background = element_rect(size = 2))

plot + theme(panel.background = element_rect(color = NA))

plot + theme(panel.background = element_rect(linetype = 'dotted'))
```

theme_blank() : draws nothing. use this element if you dont want anything drawn and no space allocated for that element. 


```{r}
plot + theme(panel.grid.minor = element_blank())

last_plot() + theme(panel.grid.major = element_blank())

last_plot() + theme(panel.background = element_blank())

last_plot() + theme(
  axis.title.x = element_blank()
)

last_plot() + theme(axis.line = element_line())
```


Customizing scales and geoms - When producing consistent theme, you may also want to tune some of the scale and geom defaults. Rather than manually specify the changes everytime you add the scale or geom. You can use the following functions to alter the default settings for scales and geoms

Scales - to change the default scale associated with an aesthetic, use set_default_scale(). 




Saving your outputs - you have two basic choices: raster or vector. Vector graphics are procedural. This means they are essentially infinitely zoomable. to the the current plot use ggsave()

ggsave() has the following arguments
- path
- three arguments that control the size: width and height for absolute size or scale to specify the size of plot relative to onscreen display
- for raster graphics the dpi argumetn controls resolution of the plot. 


Multiple plots on the same page. you will need grid. they key concept youll need to learn is view port. The default viewport takes up the entire plotting region. 

```{r}
a = qplot(date, unemploy, data = economics, geom = 'line')
b = qplot(uempmed, unemploy, data = economics) + geom_smooth(se = F)
c = qplot(uempmed, unemploy, data = economics, geom = 'path')
```


Subplots - one common layout is to have small subplot embedded drawn on top of the main plot. To achieve this effects we first plot the main plot and then draw the subplot. viewport function has parameters x,y, width, and height to control the size and position of viewport. By default the measurements are given in npc units which range from 0 to 1. the location (0,0) is bottom left (1,1) is top right. (.5,.5) is the center. YOu can also use absolute units like unit(2, cm)

```{r}
library(gridExtra)
grid.arrange(a,b,c)
```












