---
title: "Positioning"
author: "Me"
date: "7/24/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This chapter discusses position, particularly how facets are laid out on a page and how coordinate systems within a panel work. There are four components that control the position. 

Position adjustment: adjust the position of overlapping objects within a layer. these are most useful for bar and other interval geoms

Position scales: control how the values in the data are mapped to positions on the plot. Common transformations are linear and log. 

This chapter will describe the other two components and show you how all components can be used together

faceting - mechanisms for aiutomatically laying multiple plots on a page. splits data into subset and subset into a different panel on the page

coordinate systems - control how two independent position scales are combined to create a 2d coordinate system. The most common coordinate system is cartesian. 




FACETING

There are two types of faceting, facet_grid and facet_wrap. facet grid produces a 2d grid of panels defined by variables which form rows and columns. 

```{r}
library(ggplot2)
mpg2 = subset(mpg, cyl != 5 & drv %in% c('4','f'))
```



FACET GRID - lays out plots in 2d grid. you specify which variable appears on the column and which variable appears on the row

```{r}
qplot(cty, hwy, data = mpg2) + facet_grid(.~.)
```

.~ a is a single row with multiple columns. 

```{r}
qplot(cty, hwy, data = mpg2) + facet_grid(.~ cyl)
```

b ~ . single column with multiple rows. particularly useful for comparing distributions

```{r}
qplot(cty, data = mpg2, geom = 'histogram' , binwidth = 2) + facet_grid(cyl ~ .)
```


a ~ b. multiple rows and columns. youll usually want to put the variable with greatest number of levels in the column to take advantage of the aspect ratio

```{r}
qplot(cty, hwy,data = mpg2) + facet_grid(drv ~ cyl)
```


Margins - faceting a plot is like creating a contigency table. It is useful sometimes to display marginal totals as well as individual cells. you can do this with the margin argument. 

you can either specify that all margins should be displayed using margins = T or by listing names for, margins = c('sex','age').  You can also use 'grand row' or 'grand col' to produce grand row and grand column

```{r}
p = qplot(displ, hwy , data = mpg2) + geom_smooth(se = F)

p + facet_grid(cyl ~ drv, margins = T)
```



groups in the margins are controlled in the same way as groups in all other panels. the following example shows what happens when we add a colored smooth for each drive train

```{r}
qplot(displ, hwy, data = mpg2)+
  geom_smooth(aes(color = drv), method = 'lm', se = F)+
  facet_grid(cyl ~ drv, margins = T)
```



Facet wrap 

An alternative to the grid is a wrapped ribbon of plots. Instead of having a 2d grid generated by combination of two variables, facet_wrap makes a long ribbon of panels and wraps it to 2d. This is useful if you have a single variable with many levels and want to arrange the plots in more efficient manner. 



Controlling scales 

For both types of faceting you can control whether the position scales are the same in all panels(fixed) or allow to vary between panels(free). This is controlled by scales parameter

scales = 'fixed' - fixed across all panels
scales = 'free' x and y scales vary across panels
scales = 'free_x'
scales = 'free_y'

```{r}
qplot(cty, hwy, data = mpg)+
  facet_wrap(~ cyl, scales = 'free')
```



Fixed scales allow us to compare subsets on equal basis. seeing where each fits on the overall pattern. Free scales zoom on the region that each subset occupies. free scales are useful when you want to display multiple time series. 

there is an additional constraint on scales of facet_grid. all panels in the column must have the same x_scale and all panels in a row must have the same y scale. 

for facet_grid, there is an additional parameter called space which takes values 'free' or 'fixed'. 

```{r}
mpg3 = within(mpg2, {
  model  = reorder(model, cty)
  manufacturer = reorder(model, - cty)
})

models = qplot(cty, model, data = mpg3)
models + facet_grid(manufacturer ~ ., scales = 'free', 
                    space = 'free')
```
Grouping vs Faceting

faceting is an alternative to using aesthetics like color shape or size to differentiate groups. Both techniques have strengths and weaknesses based around the relative position of subsets

With faceting, each group is quite far apart in its own panel and there is no overlap between groups. This is good if the group overlaps  alot. When using aesthetics to differentiate groups, the groups are close together and may overlap but small differences are easier to see. 

```{r}
xmaj  = c(.3, .5 , 1, 3, 5)
xmin = as.vector(outer(1:10, 10^c(-1,0)))
ymaj = c(500, 1000, 5000, 10000)
ymin = as.vector(outer(1:10, 10^c(2,3,4)))

dplot = ggplot(subset(diamonds, color %in% c('D','E','G','J')),
               aes(carat, price, color = color))+
  scale_x_log10(breaks = xmaj, labels = xmaj, minor = xmin)+
  scale_y_log10(breaks = ymaj, labels = ymaj, minor = ymin)+
  scale_color_hue(limits = levels(diamonds$color)) + 
  theme(legend.position = 'none')

dplot + geom_point() + facet_grid(. ~ color)

dplot + geom_smooth(method = 'lm', se = F, fullrange = T) + facet_grid(.~color)
```

Dodging vs faceting 

Faceting can have similar effects to dodging. The main difference is labeling. the faceted plot has colored labelled above and cut below; and dodged plot has color below and cut is not explcitly labeled. in this example, the labels in the faceted plotn need some adjustment to display in a readable way. 

```{r}
qplot(color, data = diamonds, geom = 'bar', fill= cut, position = 'dodge')

qplot(cut, data = diamonds, geom = 'bar', fill = cut) + facet_grid(. ~ color)
```


Apart from labeling, the main difference between dodging and faceting occurs when two variables are nearly completely crossed, but there are some combinations that do not occur. In this case dodging becomes less useful because it only spits up the bars locally. faceting is more useful since we can control whether splitting is local

```{r}
mp4 = subset(mpg, manufacturer %in% c('audi','volkswagen','jeep'))

mp4$manufacturer = as.character(mp4$manufacturer)
mp4$model = as.character(mp4$model)

base = ggplot(mp4, aes(fill = model))+
  geom_bar(position = 'dodge')+
  theme(legend.position = 'none')

base + aes(x = model) + facet_grid(. ~ manufacturer)

last_plot() + 
  facet_grid(. ~ manufacturer, scales = 'free_x', space = 'free')

base+ aes(x = manufacturer)
```

In summary the choice between faceting and dodging are basically equivalent. 
almost crossed: faceting with shared scales ensure that all combinations are visible even if empty. this is particularly useful if missing combinations are non structural
nested: faceting with free scales allocates just enough space for each higher level group and labels each item individually


Continous variable you can facet by continous variable but you need to convert them into discrete categories first. there are three ways to do this. 

Divide the data into n bins each of the same length: cut_interval(x, n = 10) to specify the number of bins or cut_interval(length = 1) to specify the length of each interval. 

Divide the data into n bins containing apprixoximately same number of points cut_number(x, n = 10)

```{r}
mpg2$disp_ww = cut_interval(mpg2$displ, length = 1)
mpg2$disp_wn = cut_interval(mpg2$displ, n = 6)
mpg2$disp_nn = cut_number(mpg2$displ, n = 6)

plot = qplot(cty, hwy, data = mpg2) + labs(x = NULL, y = NULL)

plot + facet_wrap(~disp_ww, nrow = 1)
plot + facet_wrap(~disp_wn, nrow = 1)
plot + facet_wrap(~disp_nn, nrow = 1)
```


Coordinate systems - coordinate systems tie together two position scales to produce a 2d location. Currently ggplot2 has 7 coordinate systems. All these coordinate systems are two dimensional. Most plots use the default cartesian coordinate system, where 2d position of an element is given by combination of x and y. 

coordinate systems has two main jobs:
- combine the two position aesthetics to produce a 2d position on the plot. The position aesthetics are called x and y. 
- in coordination with the faceter, the coordinate systems draw the axes and panel backgrounds. 

Transformation - Unlike transforming the data and trasnforming the scales, transformations carried out by coordinate system change the appearance of geoms 

```{r}
library(tidyverse)
tribble(
  ~name, ~description,
  'cartesian','cartesian coordinates',
  'equal','equal scale cartesian',
  'flip','flipped cartesian',
  'trans','transformed cardesian coordinate',
  'map','map projections',
  'polar','polar coordinates'
)
```

Setting limits - coord_cartesian has arguments xlim and ylim. the key difference between coord_cartesian and scale is what appears on the plot. when setting scale_limits any data outside is thrown away, coordinate system still uses the data but only display small region

```{r}
p = qplot(disp, wt , data = mtcars) + geom_smooth()

p + scale_x_continuous(limits = c(325,500))
p + coord_cartesian(xlim = c(325,500))
```


Transformations - like limits we can also transform the data in two places at the scale level and coordinate level. coord_trans has argument x and y which should be strings naming the transformers. transforming at the coordinate system level occurs after the statistics have been computed and does affect the shape of the geom. 

```{r}
qplot(carat, price, data = diamonds, log = 'xy')+
  geom_smooth(method = 'lm')

last_plot() + coord_trans(x = 'pow10', y = 'pow10')
```

Equal scales - coord_equal ensures that x and y have equal scales. you can change the ratio argument to ensure that mapping is maintained.

Non cartesian coordinate system - 

Polar coordinates - gives rise to piechart and wind roses from bar geoms and line geoms. Polar coordinates are often used for circular data. The theta argument determines which position variable is mapped to angle. 

Stacked barchart
```{r}
pie = ggplot(mtcars, aes(x = factor(1), fill = factor(cyl))) + geom_bar(width = 1)

#pie chart
pie + coord_polar(theta = 'y')

#bulles eye polar
pie + coord_polar()
```




























