---
title: "ggplot2 notes"
author: "Me"
date: "7/26/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

```{r}
packages <- c(
  'ggplot2',
  'ggrepel',
  'patchwork',
  'ggraph',
  'dplyr',
  'gganimate',
  'gifski',
  'ggforce',
  'ggthemes',
  'maps',
  'sf',
  'tidyr',
  'concaveman',
  'remotes'
)
# Uncomment the next lines (remove the first #) to install the packages
install.packages(packages)
remotes::install_github("hrbrmstr/hrbrthemes")
# Load standard packages
library(ggplot2)
```

```{r}
ggplot(data = faithful, aes(eruptions, waiting)) + 
  geom_point()
```


```{r}
ggplot(data = faithful, aes(eruptions, waiting, color = eruptions < 3)) + 
  geom_point()
```

Setting color instead of mapping it
```{r}
ggplot(data = faithful, aes(eruptions, waiting)) + 
  geom_point(color= 'steelblue')
```
#### Exercise

Modify the code below to make the points larger squares and slightly transparent.
See `?geom_point` for more information on the point layer.

```{r}
ggplot(faithful) + 
  geom_point(aes(x = eruptions, y = waiting), size = 3,shape = 22, alpha = .5, fill = 'black')
```

Hint 1: transparency is controlled with `alpha`, and shape with `shape`
Hint 2: rememberthe difference between mapping and setting aesthetics


```{r}
ggplot(faithful)+
  geom_histogram(aes(x = eruptions))
```
### Exercise
Colour the two distributions in the histogram with different colours

```{r}
ggplot(faithful) + 
  geom_histogram(aes(x = eruptions, fill = eruptions > 3.1))
```

Hint 1: For polygons you can map two different colour-like aesthetics: `colour` 
(the colour of the stroke) and `fill` (the fill colour)


* * *

Colour the distributions in the histogram by whether `waiting` is above or below 
`60`. What happens?

```{r}
ggplot(faithful) + 
  geom_histogram(aes(x = eruptions, fill = waiting > 60))
```
Change the plot above by setting `position = 'dodge'` in `geom_histogram()` 
(while keeping the colouring by `waiting`). What do `position` control? 


```{r}
ggplot(faithful) + 
  geom_histogram(aes(x = eruptions, fill = waiting > 60), position = 'dodge')
```

```{r}
ggplot(faithful) + 
  geom_histogram(aes(x = eruptions, fill = waiting > 60), position = 'identity', alpha = .3)
```

Add a line that separates the two point distributions. See `?geom_abline` for 
how to draw straight lines from a slope and intercept.

```{r}
ggplot(faithful) + 
  geom_point(aes(x = eruptions, y = waiting))+
  geom_vline(xintercept = 3.1)
```

Layers are stacked in the order of code appearance

```{r}
ggplot(faithful, aes(eruptions, waiting))+
  geom_point()+
  geom_density_2d()
```


Statistics - Linked to geometries. geom has default stat. A layer can be created with a call to stat_*() or geom_*()

geom_bar uses stat_count by default
```{r}
library(ggplot2)
ggplot(mpg)+
  geom_bar(aes(class))
```
if you have precomputed data use identity stat
```{r}
library(tidyverse)
mpg_counted = mpg %>% count(class, name = 'count')

ggplot(mpg_counted)+
  geom_bar(aes(x = class, y = count), stat = 'identity')
```


or use geom_col for shortcut

```{r}
ggplot(mpg_counted)+
  geom_col(aes(class, count))
```

use after_stat to modify mapping from stat

```{r}
ggplot(mpg)+
  geom_bar(
    aes(x = class,
        y = after_stat(100 * count/ sum(count)))
  )+
  scale_y_continuous(labels = function(x) paste0(x, '%'))
```

Many stat provide multiple calculated values and use one by default

```{r}
ggplot(mpg)+
  geom_density(aes(x = hwy))
```

```{r}
ggplot(mpg)+
  geom_density(aes(
    x = hwy,
    y = after_stat(scaled)
  ))
```

Exercises


#### Exercises
While most people use `geom_*()` when adding layers, it is just as valid to add 
a `stat_*()` with an attached geom. Look at `geom_bar()` and figure out which
stat it uses as default. Then modify the code to use the stat directly instead
(i.e. adding `stat_*()` instead of `geom_bar()`)

```{r}
library(ggplot2)
library(tidyverse)
ggplot(mpg) + 
  stat_count(aes(x = class))+
  stat_count(aes(x = class), geom = 'point')
```

* * *

Use `stat_summary()` to add a red dot at the mean `hwy` for each group

```{r}
ggplot(mpg) + 
  geom_jitter(aes(x = class, y = hwy), width = 0.2)+
  stat_summary(aes(x = class, y = hwy), fun = mean, color = 'red', size = 1.5)


```

Hint: You will need to change the default geom of `stat_summary()`


Scales 
- used in every type of mapping. Everything inside aes() will have a scale. 
- Scales follow a predictable naming scheme scales_<aesthetic>_<type>()
- type can either be generic, continous, discrete, or binned. 


Based on the vector of class, a discrete color is picked
```{r}
ggplot(mpg)+
  geom_point(
    aes(x = displ, y = hwy, color = class)
  )
```

We take control by adding our own color explicitly. While function name is predictable, arguments are not

```{r}
ggplot(mpg)+
  geom_point(
    aes(x = displ, y = hwy, color = class)
  )+
  scale_color_brewer(type = 'qual', palette = 'Dark2', name = 'class of car')
```


X and Y are also controlled with scales

```{r}
ggplot(mpg)+
  geom_point(aes(displ, hwy))+
  scale_x_continuous(name = 'Display', breaks = c(3,5,6))+
  scale_y_continuous(name = 'Highway', trans = 'log10')
```

#### Exercises
Use `RColorBrewer::display.brewer.all()` to see all the different palettes from
Color Brewer and pick your favourite. Modify the code below to use it

```{r}
library(RColorBrewer)
display.brewer.all()

ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy, colour = class)) + 
  scale_colour_brewer(type = 'qual', palette = 'Set1')
```

* * *

Modify the code below to create a bubble chart (scatterplot with size mapped to
a continuous variable) showing `cyl` with size. Make sure that only the present 
amount of cylinders (4, 5, 6, and 8) are present in the legend.

```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy, colour = class, size = cyl)) + 
  scale_colour_brewer(type = 'qual')+
  scale_size_area(name = 'Cylinder size', breaks = c(5,6,7,8), max_size = 10)
```

Hint: The `breaks` argument in the scale is used to control which values are
present in the legend.

Explore the different types of size scales available in ggplot2. Is the default
the most appropriate here?

* * *

Modify the code below so that colour is no longer mapped to the discrete `class`
variable, but to the continuous `cty` variable. What happens to the guide?

```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy, colour = cty, size = cty))+
  guides(color = 'legend')
```
* * *


### Facets
- Use facets if you want to use small multiples
- representation of the same underlying logic
- should not be used to combine multiple separate plots
- has two facets for splitting data by categories

Faceting is often the best way to avoid overplotting
```{r}
ggplot(mpg)+
  geom_point(aes(displ, hwy))+
  facet_wrap(~class)
```

Facet grid
```{r}
ggplot(mpg)+
  geom_point(aes(x = displ, y =hwy))+
  facet_grid(year ~ drv)
```

#### Exercises
One of the great things about facets is that they share the axes between the 
different panels. Sometimes this is undiserable though, and the behaviour can
be changed with the `scales` argument. Experiment with the different possible
settings in the plot below:

```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy)) + 
  facet_wrap(~ drv, scales = 'free')
```

* * *

Usually the space occupied by each panel is equal. This can create problems when
different scales are used. Modify the code below so that the y scale differs 
between the panels in the plot. What happens?

```{r}
ggplot(mpg) + 
  geom_bar(aes(y = manufacturer)) + 
  facet_grid(class ~ ., scales = 'free_y', space = 'free')
```

Use the `space` argument in `facet_grid()` to change the plot above so each bar 
has the same width again.

* * *

Facets can be based on multiple variables by adding them together. Try to 
recreate the same panels present in the plot below by using `facet_wrap()`

```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy)) + 
  facet_wrap(~year + drv)
```

Coordinates 
- What kind of canvas should the final data be drawn on? (i.e how x and y should be interpreted)

Limits and transformation can be applied in scale or coord
- scale will apply it in the beginning
- coord will apply it in the end
- you usually want coord

- Extremely useful in cartography projection

```{r}
ggplot(mpg)+
  geom_bar(aes(x = class))+
  coord_polar(theta = 'y')+
  expand_limits(y = 70)
```

Zooming with scale removes outside limits
```{r}
ggplot(mpg)+
  geom_bar(aes(class))+
  scale_y_continuous(limits = c(0,40))
```

Zooming with coord creates a proper zoom
```{r}
ggplot(mpg)+
  geom_bar(aes(class))+
  coord_cartesian(ylim = c(0,40))
```

#### Exercises
In the same way as limits can be set in both the positional scale and the coord,
so can transformations, using `coord_trans()`. Modify the code below to apply a
log transformation to the y axis; first using `scale_y_continuous()`, 
and then using `coord_trans()`. Compare the results — how do they differ?

Better because break and break points are better.
```{r}
ggplot(mpg) + 
  geom_point(aes(x = hwy, y = displ))+
  coord_trans(y = 'log10')
```







