---
title: "Workshop pt2"
author: "Me"
date: "7/27/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

```{r, echo=FALSE}
packages <- c(
  'ggplot2',
  'ggrepel',
  'patchwork',
  'ggraph',
  'dplyr',
  'gganimate',
  'gifski',
  'ggforce',
  'ggthemes',
  'maps',
  'sf',
  'tidyr',
  'concaveman',
  'remotes'
)
# Uncomment the next lines (remove the first #) to install the packages
install.packages(packages)
remotes::install_github("hrbrmstr/hrbrthemes")
# Load standard packages
library(ggplot2)
```

```{r}
library(ggplot2)
```

### Theme
Theming defines the feel and look of your final visualisation and is something
you will normally defer to the final polishing of the plot. It is very easy to 
change looks with a prebuild the

```{r}
ggplot(mpg) + 
  geom_bar(aes(y = class)) + 
  facet_wrap(~year) + 
  theme_minimal()
```


```{r}

library(ggplot2)
library(extrafont)
loadfonts(device = "win")

ggplot(mpg) + 
  geom_bar(aes(y = class)) + 
  facet_wrap(~year) + 
  labs(title = "Number of car models per class",
       caption = "source: http://fueleconomy.gov",
       x = NULL,
       y = NULL) +
  scale_x_continuous(expand = c(0, NA)) + 
  theme_minimal() + 
  theme(
    text = element_text('Segoe Print'),
    strip.text = element_text(face = 'bold', hjust = 0),
    plot.caption = element_text(face = 'italic'),
    panel.grid.major = element_line('white', size = 0.5),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.ontop = TRUE
  )
```


## Extensions

Why extension ? 
- many extension are niche specific and better developed by experts in the field. 
- easier to promote focused packages


```{r}
install.packages('ggdist')
install.packages('ggcharts')
install.packages('ggthemes')
library(ggdist)
library(ggcharts)
library(ggthemes)
```

Plot composition
- plot layout should be flexible and well aligned
- Many approaches
 - facet hacking
 - gridExtra::grid.arrange()
 - ggpubr::ggarrange()
 - cowplot::plot_grid()
 - patchwork



```{r}
p1 = ggplot(msleep)+
  geom_boxplot(aes(x = sleep_total, y = vore, fill = vore))

p2 = ggplot(msleep)+
  geom_bar(aes(y = vore, fill = vore))

p3 = ggplot(msleep)+
  geom_point(aes(x = bodywt, y = sleep_total, color = vore))+
  scale_x_log10()

(p1 | p2) / p3
```

```{r}
library(patchwork)
p_all = (p1 | p2) / p3

p_all +
  plot_layout(guides = 'collect')
```

& add ggplot2 elements to all subplots
```{r}
p_all & 
  theme(legend.position = 'None')
```

```{r}
p_all + plot_annotation(
  title = 'Mammalian Sleep Patterns',
  tag_levels = 'A'
)
```

#### Excercises
Patchwork will assign the same amount of space to each plot by default, but this
can be controlled with the `widths` and `heights` argument in `plot_layout()`. 
This can take a numeric vector giving their relative sizes (e.g. `c(2, 1)` will 
make the first plot twice as big as the second). Modify the code below so that
the middle plot takes up half of the total space:

```{r}
p <- ggplot(mtcars) + 
  geom_point(aes(x = disp, y = mpg))

p_all = p + p + p

p_all + plot_layout(widths = c(1,.5,1))

p_all + plot_layout(widths = unit(c(5,1,1), c('cm','null','null')))
```

* * *

The `&` operator can be used with any type of ggplot2 object, not just themes.
Modify the code below so the two plots share the same y-axis (same limits)

```{r}
p1 <- ggplot(mtcars[mtcars$gear == 3,]) + 
  geom_point(aes(x = disp, y = mpg))
p2 <- ggplot(mtcars[mtcars$gear == 4,]) + 
  geom_point(aes(x = disp, y = mpg))
p_all = p1 + p2

p_all &
  scale_y_continuous(limits = c(10,40))
```

* * *

Patchwork contains many features for fine tuning the layout and annotation. Very
complex layouts can be obtained by providing a design specification to the 
`design` argument in `plot_layout()`. The design can be defined as a textual 
representation of the cells. Use the layout given below. How should the textual 
representation be undertood.

```{r}
p1 <- ggplot(mtcars) + 
  geom_point(aes(x = disp, y = mpg))
p2 <- ggplot(mtcars) + 
  geom_bar(aes(x = factor(gear)))
p3 <- ggplot(mtcars) + 
  geom_boxplot(aes(x = factor(gear), y = mpg))

layout <- '
11#
#22
3##
'
p1 + p2 + p3 + plot_layout(design = layout)

area(1, 1)
```


### Animations

- Story telling and attention grabbing
- Our eyes are drawn to movement and our mind think in motion
- Many way to make animations. gganimate does it grammary way

```{r}
library(gganimate)
ggplot(economics)+
  geom_line(aes(x = date, y = unemploy))+
  transition_reveal(along = date)
```


```{r}
ggplot(mpg)+
  geom_bar(aes(x = factor(cyl)))+
  labs(title = 'Number of cars in {closest_state} by number of cylinders')+
  transition_states(states = year)+
  enter_grow()+
  exit_fade()
```

#### Exercises
The animation below will animate between points showing cars with different 
cylinders.

```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy, group = seq_len(nrow(mpg)))) + 
  ggtitle("Cars with {closest_state} cylinders") + 
  transition_states(factor(cyl))
```

gganimate uses the `group` aesthetic to match observations between states. By 
default the group aesthetic is set to the same value, so observations are 
matched by their position (first row of 4 cyl is matched to first row of 5 cyl 
etc.). This is clearly wrong here (why?). Add a mapping to the `group` aesthetic
to ensure that points do not move between the different states.

* * *

In the presence of discrete aesthetic mappings (`colour` below), the group is
deduced if not given. The default behaviour of objects that appear and disappear
during the animation is to simply pop in and out of existance. `enter_*()` and
`exit_*()` functions can be used to control this behaviour. Experiment with the
different enter and exit functions provided by gganimate below. What happens if 
you add multiple enter or exit functions to the same animation?

```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy, colour = factor(cyl))) + 
  ggtitle("Cars with {closest_state} cylinders") + 
  transition_states(factor(cyl))
```

* * *

In the animation below (as in all the other animations) the changes happens at
constant speed. How values change during an animation is called easing and can
be controlled using the `ease_aes()` function. Read the documentation for 
`ease_aes()` and experiment with different easings in the animation.

```{r}
mpg2 <- tidyr::pivot_longer(mpg, c(cty,hwy))
ggplot(mpg2) + 
  geom_point(aes(x = displ, y = value)) + 
  ggtitle("{if (closest_state == 'cty') 'Efficiency in city' else 'Efficiency on highway'}") + 
  transition_states(name)+
  ease_aes(y = 'cubic-in-out')
  
```

ANNOTATION
- you cant tell a story without words
- this has often been relegated to post processing
- Recent improvements means most can be done with codes
- two key packages: ggrepel, ggforce


```{r}
ggplot(mtcars, aes(x = disp, y = mpg))+
  geom_point()+
  geom_text(aes(label = row.names(mtcars)))
```

Can be fixed with ggrepel
```{r}
library(ggrepel)

ggplot(mtcars, aes(x = disp, y = mpg))+
  geom_point()+
  geom_text_repel(aes(label = row.names(mtcars)))
```

```{r}
library(ggforce)

ggplot(mtcars, aes(disp, mpg))+
  geom_point()+
  geom_mark_ellipse(
    aes(filter = gear == 4,
        label = '4 gear cars',
        description = 'Cars with fewer gears
        tend to have both higher yield and lower displacement')
  )
```

#### Exercises
ggrepel has a tonne of settings for controlling how text labels move. Often, 
though, the most effective is simply to not label everything. There are two
strategies for that: Either only use a subset of the data for the repel layer,
or setting the label to `""` for those you don't want to plot. Try both in the 
plot below where you only label 10 random points.

```{r}
mtcars2 <- mtcars
mtcars2$label <- rownames(mtcars2)
points_to_label <- sample(nrow(mtcars), 10)
ggplot(mtcars2, aes(x = disp, y = mpg)) + 
  geom_point() + 
  geom_text_repel(aes(label = label), data = mtcars2[points_to_label,])
```

* * *

Explore the documentation for `geom_text_repel`. Find a way to ensure that the 
labels in the plot below only repels in the vertical direction

```{r}
mtcars2$label <- ""
mtcars2$label[1:10] <- rownames(mtcars2)[1:10]
ggplot(mtcars2, aes(x = disp, y = mpg)) + 
  geom_point() + 
  geom_text_repel(aes(label = label), direction = 'y', vjust = 1)
```

* * *

ggforce comes with 4 different types of mark geoms. Try them all out in the code
below:

```{r}
ggplot(mtcars, aes(x = disp, y = mpg)) +
  geom_point() + 
  geom_mark_circle(aes(filter = gear == 4,
                        label = '4 gear cars'), radius = F)+
  coord_cartesian(clip = 'off')
```

```{r}
ggplot(mtcars, aes(x = disp, y = mpg)) +
  geom_point() + 
  geom_mark_hull(aes(fill = factor(gear), labels = gear))
```

```{r}
ggplot(mtcars, aes(x = disp, y = mpg)) +
  geom_point() + 
  geom_mark_hull(aes(group = factor(gear), labels = gear), fill = NA, color = NA)
```


Networks 
Multiple packages tried to add network capabilities to ggplot2
- GGally
- ggnetwork
- geomnet
- ggtree
- ggdag
- ggraph


```{r}
library(ggraph)
library(tidygraph)
library(tidyverse)

graph = create_notable('zachary') %>%
  mutate(clique = as.factor(group_infomap()))

ggraph(graph)+
  geom_edge_link()+
  geom_node_point(aes(color = clique), size = 2)
```


```{r}
ggraph(graph)+
  geom_mark_hull(aes(x, y, fill = clique))+
  geom_edge_link()+
  geom_node_point( size = 2)
```

ggraph understands objects represented as networks
```{r}
iris_clust = hclust(dist(iris[,1:4]))

ggraph(iris_clust)+
  geom_edge_bend()+
  geom_node_point(aes(filter = leaf))
```


#### Exercies
Most network plots are defined by a layout algorithm, which takes the network 
structure and calculate a position for each node. The layout algorithm is 
global and set in the `ggraph()`. The default `auto` layout will inspect the
network object and try to choose a sensible layout for it (e.g. dendrogram for a
hierarchical clustering as above). There is, however no optimal layout and it is
often a good idea to try out different layouts. Try out different layouts in the
graph below. See the [the website](https://ggraph.data-imaginist.com/reference/index.html)
for an overview of the different layouts.

```{r}
ggraph(graph, layout = 'gem') + 
  geom_edge_link() + 
  geom_node_point(aes(colour = clique), size = 3)
```

* * *

There are many different ways to draw edges. Try to use `geom_edge_parallel()` 
in the graph below to show the presence of multiple edges

```{r}
highschool_gr <- as_tbl_graph(highschool)
ggraph(highschool_gr) + 
  geom_edge_parallel() + 
  geom_node_point()
```

Faceting works in ggraph as it does in ggplot2, but you must choose to facet by
either nodes or edges. Modify the graph below to facet the edges by the `year` 
variable (using `facet_edges()`)

```{r}
ggraph(highschool_gr) + 
  geom_edge_fan() + 
  geom_node_point()+
  facet_edges(~year)
```

Looks 
- theming is one of the easiest way to extend
-lots of options
 - ggthemes
 - tvthemes
 - ggtech
 - ggthemr
 - hrbrthemes
 
- often coupled with color scales
- dont go too crazy unless wanted


```{r}
install.packages('tvthemes')
install.packages('ggthemr')
library(hrbrthemes)
library(tvthemes)


p <- ggplot(mtcars, aes(mpg, wt)) +
  geom_point(aes(color = factor(carb))) +
  labs(
    x = 'Fuel efficiency (mpg)', 
    y = 'Weight (tons)',
    title = 'Seminal ggplot2 example',
    subtitle = 'A plot to show off different themes',
    caption = 'Source: It’s mtcars — everyone uses it'
  )

p + theme_ipsum()

p + scale_color_excel() + theme_excel()

p + theme_excel()
```

Drawing Anything

If all else fails, everything is just points, lines, polygons, and text. If faceting fails, use patchwork. 

```{r}
states = c(
  'eaten',
  'eaten but you said you didn\'t',
  'cat took it',
  'for tonight',
  'will decompose slowly'
)

pie = data.frame(
  state = factor(states, levels = states),
  amount = c(4,3,1,1.5,6),
  stringsAsFactors = F
)

ggplot(pie)+
  geom_col(aes(x = 0, y = amount, fill = state))+
  coord_polar(theta = 'y')
```

```{r}
ggplot(pie)+
  geom_col(aes(x = 0, y = amount, fill = state))+
  coord_polar(theta = 'y')+
  scale_fill_tableau(
    name = NULL,
    guide = guide_legend(ncol = 2)
  )+
  theme_void()+
  theme(legend.position = 'top',
        legend.justification = 'left')
```


```{r}
ggplot(pie)+
  geom_arc_bar(aes(
    x0 = 0,
    y0 = 0,
    r0 = 0,
    r = 1,
    amount =amount,
    fill = state
  ), stat = 'pie')+
  coord_fixed()+
  scale_fill_tableau(name = NULL, guide = guide_legend(ncol = 2))+
  theme_void()+
  theme(legend.position = 'top',
        legend.background = element_rect(color = 'black', fill = 'gray', linetype = 3))
```


```{r}
ggplot(mpg)+
  geom_bar(aes(x = hwy), stat = 'bin')

ggplot(mpg)+
  geom_bar(aes(hwy))+
  scale_x_binned(n.breaks = 30, guide = guide_axis(n.dodge = 2))+
  theme_classic()
```



















